Visualizing demographic evolution using geographically inconsistent census data

### 2019-08-24 paper/intro.tex
from overleaf

Yet new challenges have also emerged, especially at the convergence of research
on neighbourhood effects and neighbourhood dynamics. Neighbourhood effects
research assumes knowledge about the nature and scope of "the neighbourhood"
that presumably shapes individual outcomes~\citep{Kwan2018,Shelton2019}.
Concurrently, researchers note that neighbourhoods are not necessarily fixed
containers in which other processes occur, but themselves dynamically
evolve~\citep{Delmelle2017,Reades2019,li2018new}. \revision{In this work, we
consider neighbourhoods as \emph{formal regions}~\citep{montello2003regions},
geographically continous areas with similar data characteristics, which these
advances made more tractable to approach in a data-driven fashion.} The result
is to open up key assumptions about neighbourhoods for theoretical and empirical
examination: how do we appropriately define and compare neighbourhoods at a
given time?; how do we appropriately define and compare the temporal
trajectories of neighbourhoods?; and can we do both at once, "fully
interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
they came from and where they are going?

In principle, much of the recent research is committed to the proposition that
neighbourhoods are open and evolving entities. Ironically, its empirical
practice tends to rely on methods that require fixed geographical regions. This
requirement is difficult to satisfy, as most longitudinal datasets are based on
pre-defined tabulation areas that are routinely modified by data collection
agencies, usually to follow population changes.

The standard approach then is to \emph{geographically harmonise} the data. This
involves interpolating existing measurements into a common set of
regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent computational tools
have somewhat simplified this process~\citep{rey2018spatio}, but it still
involves non-trivial questions: which geometry to use as target, how to
apportion the variables, or how to combine data from different sources. Further,
these question do not necessarily have optimal answers. Indeed, regardless of
how well this process is performed, it still introduces
errors~\citep{Logan2016}, even when additional data is
provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
\emph{artificial data points} that can potentially lead to inaccurate results,
even though they are seldom interpreted as such. Nevertheless, because there has
been no viable alternative, and the results often appear plausible, these
concerns are generally overlooked. The result is that the harmonisation approach
is virtually mandatory in the current literature: \emph{"(...) tract-by-tract
comparison is not possible unless data from 2000 is interpolated to 2010
boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
comparison since data are not representative of the same spatial units. (...)"}
~\citep{Allen2018}.

The main contribution of this paper is a method for longitudinal data processing
that works with the original data by leveraging a network based representation.
It enables tract-by-tract comparison and the identification of patterns of
demographic evolution \emph{without geographic harmonisation}.

@@

Real world measurements are seldom as homogeneous and noise-free as this
artificial example. By splitting and merging the data to fit arbitrary borders,
that were not necessarily coherent at the time of the measurement, harmonisation
increases the distance between measurement and reality.

@@

Instead, we propose a network-based representation. A \emph{network} (also
called a \emph{graph}) is a collection of entities (nodes) that are related to
each other (edges). In this case, each different aggregation area is represented
as a node and we connect nodes that have overlapping geographical areas\revision{ in
different times or are neighbours in the same time}, leading to the network
illustrated on the right of Figure~\ref{fig:intuition}. By partitioning the
network into connected nodes that are similar, we are effectively identifying
clusters in the spatio-temporal data, as illustrated by the colours of the nodes
on the right side of Figure~\ref{fig:intuition}. Further, all the possible paths
of change can be obtained by computing sequences of nodes over time, in this
case: (A, C, F), (A, D, F), (A, D, G), and (B, E, H). This representation is
also suited for geographically consistent regions, as illustrated by the stable
park in this example, and is therefore a generalisation of the traditional
paradigm.

@@

In the following, we demonstrate that this network representation allows us to
study neighbourhood change in a way that does not require prior interpolation.

>>>>

Yet new challenges have also emerged, especially at the convergence of research
on neighbourhood effects and neighbourhood dynamics. Neighbourhood effects
research assumes knowledge about the nature and scope of "the neighbourhood"
that presumably shapes individual outcomes~\citep{Kwan2018,Shelton2019}.
Concurrently, researchers note that neighbourhoods are not necessarily fixed
containers in which other processes occur, but themselves dynamically
evolve~\citep{Delmelle2017,Reades2019,li2018new}.  The result is to open up key
assumptions about neighbourhoods for theoretical and empirical examination: how
do we appropriately define and compare neighbourhoods at a given time?; how do
we appropriately define and compare the temporal trajectories of
neighbourhoods?; and can we do both at once, "fully
interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
they came from and where they are going?

In principle, much of the recent research is committed to the proposition that
neighbourhoods are open and evolving entities. Ironically, its empirical
practice tends to rely on methods that require fixed geographical regions. This
requirement is not trivial to satisfy in general, as most longitudinal datasets
are based on pre-defined tabulation areas that are routinely modified by data
collection agencies, usually to follow population changes.

The standard approach then is to \emph{geographically harmonise} the data. This
involves interpolating existing measurements into a common set of
regions~\citep{Logan2014,Hallisey2017,Allen2018}. \revision{1.18}{Recent
computational tools have somewhat simplified this
process~\citep{rey2018spatio}}, but it still involves non-trivial questions:
which geometry to use as target, how to apportion the variables, or how to
combine data from different sources. Further, these question do not necessarily
have optimal answers. Indeed, regardless of how well this process is performed,
it still introduces errors~\citep{Logan2016}, even when additional data is
provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
\emph{artificial data points} that can potentially lead to inaccurate results,
even though they are seldom interpreted as such. Nevertheless, because there has
been no viable alternative, and the results often appear plausible, these
concerns are overlooked. The result is that the harmonisation approach is
virtually mandatory in the current literature: \emph{"(...) tract-by-tract
comparison is not possible unless data from 2000 is interpolated to 2010
boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
comparison since data are not representative of the same spatial units. (...)"}
~\citep{Allen2018}.

\unsure{CHECK blue text below}
\revision{1.1}{In some cases, harmonisation might seem methodologically
adequate, if a study is focused on a small subregion, fixed in time, limiting
and modifying the data to fit that specific region of interest sounds not only
reasonable, but recommended, albeit the creation of artificial data points.
However, that is a simplistic view that ignores interactions beyond those
borders: each different portion of the area of interest may belong to larger
homogeneous processes beyond these arbitrarily defined borders. This effect,
along with the creation of artificial points, is caused by the geometry mismatch
between the data and the region of interest. Instead of forcing one set of
borders into another, our method sidesteps this issue by changing how the data
is represented and interpreted.

The main contribution of this paper is a method for longitudinal geographical
data representation and processing that works with the original data by
leveraging a network based representation. It enables direct comparison and
temporal regionalisation without the need to match arbitrarily defined borders.
Even in the case of a small, temporally fixed, region of interest, our approach
leads to a better understanding of its dynamics by exploring the related
data-driven homogeneous regions. }

@@

\revision{2.6.4}{While didactic, this example is not realistic because each
building is visible in Figure~\ref{fig:intuition}, so we know each individual of
the tabulation area, which is never the case. Further, real measurements are
seldom as homogeneous and noise-free as this artificial example.  In this
example, one possible harmonisation would split region A  in two to provide
different ancestors to C ($A_C$) and D ($A_D$). Then $A_D$ would not be
considered when analysing the evolution of C, nor $A_C$ for $D$. This is
inaccurate, because $A$ was the original state for both evolutions, and we
should not assume that these processes are disconnected.} Either way, by
splitting and merging the data to fit arbitrary borders, the harmonisation
process increases the distance between the data and reality by creating
artificial data points.

@@

Instead, we propose a network-based representation. A \emph{network} (also
called a \emph{graph}) is a collection of entities (nodes) that are related to
each other (edges). In this case, each different aggregation area is represented
as a node and we connect nodes that have overlapping geographical areas\revision{1.2}{ in
different times or are neighbours in the same time}, leading to the network
illustrated on the right of Figure~\ref{fig:intuition}. By partitioning the
network into connected nodes that are similar, we are effectively identifying
clusters in the spatio-temporal data, as illustrated by the colours of the nodes
on the right side of Figure~\ref{fig:intuition}. Further, all the possible paths
of change can be obtained by computing sequences of nodes over time, in this
case: (A, C, F), (A, D, F), (A, D, G), and (B, E, H). This representation is
also suited for geographically consistent regions, as illustrated by the stable
park in this example, and is therefore a generalisation of the traditional
paradigm.

@@

### 2019-08-24 paper/method.tex
from overleaf

\begin{figure}
    \centering 
    \includegraphics[width=0.975\linewidth]{overview.pdf}
    \caption{Overview of the proposed method. A network is generated combining the
        original census data, encoding the changing geographical information.
        The network is partitioned into an hierarchy~\citep{markus2017}. The
        characteristics and evolution of the clusters are then visually
        represented.
        \label{fig:overview}}
\end{figure}

@@

Each area of each census year is represented as a node, and edges are placed
between nodes if the corresponding regions share geographic borders in the same
year. Further, edges are placed between nodes if the corresponding regions
belong to sequential years and \revision{their geometries intersect. To avoid
spurious connections caused by geometry fluctuations, one of the geometries is
slightly shrunk before the intersection, using a buffer of -1e6. More
importantly, while weights will be associated with these edges before they are
processed, they are not derived from the geometry, but from the data. The actual
intersection area is not considered in this representation. } This approach
leads to a single network representing the whole spatio-temporal space of the
data. Our objective then becomes to identify partitions of this network such
that the nodes of each partition are more similar between themselves than to the
other nodes.

\subsection{Geographic content clustering}
Having tied the regions together into a network, we can now partition it to
identify similar sets of regions. \revision{In essence, we are performing
temporal regionalisation by applying a regionalisation method over this
spatio-temporal network.}

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017} with
median linkage. The initial watershed step is performed to create an initial
clustering and reduce the running time of the SMM. We introduced one new
parameter to this method: a maximum distance threshold for the merges, to avoid
the early merging of outliers. We refer the reader to the original
paper~\citep{markus2017} for more details, including a complete performance
evaluation using several metrics. \revision{We chose this algorithm because it
is fast, simple, and easily customisable, but our methodology should work with
any hierarchical network clustering algorithm.}

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes.
\revision{The regions connected by those edges will be merged on the first
stages of the clustering, since the nodes they connect are, by definition, as
similar as possible\footnote{This assertion relies on the euclidean distance and
its relationship with the space where k-nn operates.}, leaving the remaining
steps of the hierarchy to be determined only by the geographical edges. We
experimented with different numbers of augmentation edges, but the results were
not consistent, since the distribution of the edges is data dependent. Adding
two edges per node was the smallest number of edges that led to stable and
consistent results in the scenarios available in our prototype. Since the
problem of balancing the data space with the geographical space is relevant for
geographical data analysis, this idea potentially warrants further exploration,
beyond the scope of this work.}

@@

We also use these statistical measurements to discover what characterises each
cluster, that is, what makes it different from the others.  We define the
\emph{relevance} of a variable of an aspect based on the distance between the
interquartile ranges (IQR) of the clusters in the same hierarchical level. If
the IQRs overlap for all clusters, that variable is not relevant to the
characterisation of the cluster, but if the IQRs are distant, it means that this
specific range of values is something that only occurs in this cluster.
Examining IQRs therefore provides users a straightforward visual method for
determining what variables most clearly define a given cluster.

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what groups exist and where they are, we are also interested in the
evolution of these groups. To examine this process of evolution directly, we
introduce the concepts\revision{ of \emph{temporal paths} and \emph{trajectories}.

We call a temporal path any sequence of nodes in our representation network such
that the temporal information associated with the nodes only increases. For
instance, in Figure~\ref{fig:intuition}, the sequences ACF, ADF, ADG, and BEH
are temporal paths. With harmonised data, the time-series of to each region
would form a temporal path, each node would be connected only to its older and
newer versions, belonging to only one temporal path, as illustrated by the path
BEF. Since our data is not harmonised, more connections are allowed and each
node can belong to an arbitrary number of paths.

Semantically, this is a generalisation of the idea of geographical time-series,
because each temporal path is one possible option for the data to change over
time. Returning to Figure~\ref{fig:intuition}, the paths ACF, ADF, and ADG all
start on the same homogeneous region, but evolve differently over time. In other
words, this network encodes the information that portions of the region A
changed to form regions C and D, but we do not know specifically which parts,
nor we need to, since the same region can belong to several temporal paths.
Interestingly, when interpreted in this framework, geographical harmonisation is
a method to split and/or merge nodes so each belongs to a single temporal path.

@@

\subsection{User interface}
\label{sec:ui}
To validate and explore the results of our methodology, we built a user
interface, illustrated in Figure~\ref{fig:ui}, considering census tract (CT)
level data from the Chicago region between 1970 and 2010. This region is known
for its entrenched racial divide and the emergence of a \emph{'young urban'}
population with a higher education level~\citep{Delmelle2016,Delmelle2017}. More
details are presented in Section~\ref{sec:study}.

\begin{figure*}
    \centering 
    \includegraphics[width=0.975\linewidth]{ui.pdf}
    \caption{Initial interface of our method showing the demographic evolution of Chicago. 
        \textbf{(a)}: Configuration panel with the current clustering parameters and controls.
        \textbf{(b)}: Cluster overview illustrating the most relevant aspect for each cluster. 
        \textbf{(c)}: Trajectories overview and the general evolution of the population, geographical information, and how it changed. 
        \textbf{(d)}: Details of the selected trajectories, including precise geographic locations, population numbers, and the composition of the aspects.\label{fig:ui}}
\end{figure*}

As illustrated by Figure~\ref{fig:ui}, our proposed interface heavily relies on
colour to express cluster-related information. We adopted this convention
because colours can be used in all our visual tools in a coherent manner.
However, there is a limit on the number of distinct colours that can be used. We
limited the number of clusters to eight because this was the largest number of
colours that we could reliably and accessibly use, derived from the 8-class
Dark2 set from ColorBrewer~\citep{ColorBrewer}.

The configuration panel, on top left in Figure~\ref{fig:ui}, displays which
aspects were used and their weights (following Equation~\ref{eq:dist}). It also
includes other configuration options that can be altered without re-processing
the data, such as the number of clusters and the colour option. The gear button
allows access to the other configuration options that do require further
processing, such as changing location, aspects, and weights.

The cluster overview panel, on the bottom left in Figure~\ref{fig:ui}, displays
a brief summary of each cluster, based on the distance between the IQRs, as
detailed in Section~\ref{sec:relevance}. The \emph{View all} button opens a new
panel where all aspects are included, while the chevron at the side lets the
user expand each cluster separately.

We adopted an \emph{enhanced} version of the traditional boxplot, which includes
the IQRs for the other clusters, in slightly larger and faded black rectangles.
We also colour the current IQR according to its relevance.  For instance, the
boxplot that summarises the purple cluster illustrated in Figure~\ref{fig:ui},
detailed on Figure~\ref{fig:boxplot}, illustrates that this cluster is best
defined by the proportion of the population with four or more years of college.
The user can quickly see that this is relevant because the corresponding IQR is
coloured with the highest relevance present in the legend. It is also clear
that, while this cluster includes CTs that have between 10\% to 90\% of people
in this variable, approximately, half of them have about 60\% of the population
with four or more years of college. Since all the other IQRs are well separated,
this is a defining characteristic of this cluster. Conversely, the proportion of
the population with one to three years of college is not relevant, as indicated
by black fill in the rectangle representing the IQR of this cluster, in
overlapping position with the rectangles of the other clusters. By clicking on
the coloured bar above the boxplot, the user can select all trajectories that
contain this cluster at any point in time.

@@

The trajectories overview aims to convey basic information about the
trajectories, where they are, and what changes are involved. This is done using
three sub panels. The first, on the left, contains a Sankey diagram illustrating
the evolution of the clusters over time. The widths are proportional to the
population involved. In our example in Figure~\ref{fig:ui}, the orange and green
clusters contain most of the population and are fairly stable over time. The
pink cluster is small and mostly stable. The purple cluster is increasing,
mostly by incorporating areas that were previously orange. Since the purple
group corresponds to the emergent 'young urban' group, this corroborates the
findings of Delmelle~\citep{Delmelle2016,Delmelle2017}, showing that our
network-based method can recover results from the traditional data processing
approach.

In the next panel, illustrated in the top middle of Figure~\ref{fig:ui}, is a
transition matrix between the clusters. It indicates a rounded percentage of the
population whose area changed between each pair of clusters. This kind of table
can be found in the related literature~\citep{Delmelle2016}, so it is familiar
to the advanced users. It not only informs the proportional changes, but allows
the selection of the corresponding trajectories for further analysis.

The panel in the top right of Figure~\ref{fig:ui} is a map of the region under
analysis, summarising the geographical evolution of the clusters over time.  The
colours are derived from the clusters involved in each trajectory, which are
consistent across the linked views.

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, as illustrated in
Figure~\ref{fig:ui}. This panel contains two main regions: the small multiple
maps, depicting the clusters at each year, and the stacked bar plots that
summarise the overall composition of these regions. In this example, the maps
show the transition from orange to green and purple in several regions over
time. Clicking on a region in these maps will bring up a new panel with the
original census data of this specific region. The actual population numbers are
below the maps.

Each aspect is represented by a stacked bar plot, where the width of each
rectangle corresponds to \revision{the percentage of population in that variable
over the considered period. In this case, a little less than half of the
population are married, and the percentage that are Widowers or Divorced is
roughly similar. About half of the population work in Administrative jobs, a
third never completed high-school, approximately fourty percentq have gross
family income below 50,000USD per year. About sixty percent identify as white.
Placing the mouse over one of the bars will open a small panel with the temporal
evolution of the population percentage of that specific variable, and clicking
on the chevron on the right side expands the corresponding aspect, showing
census level statistics, with details of the temporal evolution of each variable
and also the corresponding IQRs for the whole city. Those values are potentially
different, for instance, while the overall population of this region is about
sixty percent White, it includes regions with wildly different percentages, so
the average percentage of White population over the CTs is a little over fourty
percent.}

>>>>

\begin{figure}
    \centering 
    \includegraphics[width=0.975\linewidth]{overview.pdf}
    \caption{Overview of the proposed method. 
        \textbf{(a)} A network is generated from the original data,
        encoding the changing geographical information. 
        \textbf{(b)} The network is partitioned into an hierarchy. 
        \textbf{(c)} The characteristics and evolution of the clusters are then
        visually represented.
        \label{fig:overview}}
\end{figure}

@@

Each area of each census year is represented as a node, and edges are placed
between nodes if the corresponding regions share geographic borders in the same
year. Further, edges are placed between nodes if the corresponding regions
belong to sequential years and \revision{1.6/2.6.3}{their geometries intersect.
To avoid spurious connections caused by small fluctuations, one of the
geometries is slightly shrunk before the intersection, using a buffer of -1e6.
While the result would not change considerably by adding more connections
between nearby regions, these extra connections would greatly increase the
computational cost. More importantly, while weights will be associated with
these edges before they are processed, they are not derived from the geometry,
but from the data. The actual intersection area is not considered in this
representation. } This approach leads to a single network representing the whole
spatio-temporal space of the data.

\subsection{Geographic content clustering}
Having tied the regions together into a network, we can now partition it to
identify similar sets of regions. \revision{2.3}{In essence, we are performing
temporal regionalisation by applying a clustering method over this
spatio-temporal network.}

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017}. The
initial watershed step is performed to create an initial clustering and reduce
the running time of the SMM. We introduced one new parameter to this method: a
maximum distance threshold for the merges, to avoid the early merging of
outliers. We refer the reader to the original paper~\citep{markus2017} for more
details, including a complete performance evaluation using several metrics.
\revision{1.5/2.2}{We chose this algorithm because it is fast,
simple, and easily customisable, but our methodology should work with any
hierarchical network clustering algorithm. However, we caution against the use
of single-linkage methods due to their tendency to form chain-like clusters with
elements that are similar pairwise, but significantly different on the ends.
That is also the reason we did not use a Watershed hierarchy
directly~\citep{najman2011equivalence}.}

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes.
\revision{2.4}{The regions connected by those edges will be merged on the first
stages of the clustering, since the nodes they connect are, by definition, as
similar as possible\footnote{This assertion relies on the euclidean distance and
its relationship with the space where k-nn operates.}, leaving the remaining
steps of the hierarchy to be determined only by the geographical edges. We
experimented with different numbers of augmentation edges, but the results were
not consistent, since the distribution of the edges is data dependent. Adding
two edges per node was the smallest number of edges that led to stable and
consistent results in the scenarios available in our prototype. Since the
problem of balancing the data space with the geographical space is relevant for
geographical data analysis, this idea potentially warrants further exploration,
beyond the scope of this work.}

@@

We also use these statistical measurements to discover what characterises each
cluster, that is, what makes it different from the others.  We define the
\emph{relevance} of a variable of an aspect based on the distance between the
\revision{1.8}{interquartile} ranges (IQR) of the clusters in the same hierarchical level. If
the IQRs overlap for all clusters, that variable is not relevant to the
characterisation of the cluster, but if the IQRs are distant, it means that this
specific range of values is something that only occurs in this cluster.
Examining IQRs therefore provides users a straightforward visual method for
determining what variables most clearly define a given cluster.

To allow for an easier visualisation of these ideas, we represent it using an
\emph{enhanced} version of the traditional boxplot, which includes the IQRs for
the other clusters, in slightly larger and faded black rectangles. We also
colour the current IQR according to its relevance, as illustrated in
Figure~\ref{fig:boxplot}. In this example, this cluster is best defined by the
proportion of the population with four or more years of college. The user can
quickly see that this is relevant because the corresponding IQR is coloured with
the highest relevance present in the legend. It is also clear that, while this
cluster includes CTs that have between 10\% to 90\% of people in this variable,
approximately, half of them have about 60\% of the population with four or more
years of college. Since all the other IQRs are well separated, this is a
defining characteristic of this cluster. Conversely, the proportion of the
population with one to three years of college is not relevant, as indicated by
black fill in the rectangle representing the IQR of this cluster, in overlapping
position with the rectangles of the other clusters.

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what groups exist and where they are, we are also interested in the
evolution of these groups. To examine this process of evolution directly, we
introduce the concepts\revision{2.6.4}{ of \emph{temporal paths} and \emph{trajectories}.

We call a temporal path any sequence of nodes in our representation network such
that the temporal information associated with the nodes only increases. For
instance, in Figure~\ref{fig:intuition}, the sequences ACF, ADF, ADG, and BEH
are temporal paths. With harmonised data, as illustrated by the path BEF, the
time-series of each region would form a temporal path, each node would be
connected only to its older and newer versions, and would belong to only one
temporal path. Since our data is not harmonised, more connections are allowed
and each node can belong to an arbitrary number of paths.

Semantically, this is a generalisation of the idea of geographical time-series,
because each temporal path is one possible option for the data to change over
time. Returning to Figure~\ref{fig:intuition}, the paths ACF, ADF, and ADG all
start on the same homogeneous region, but evolve differently over time. In other
words, this network encodes the information that portions of the region A
changed to form regions C and D, but we do not know specifically which parts.
Nor do we need to, since the same region can belong to several temporal paths.
Interestingly, when interpreted in this framework, geographical harmonisation is
a method to split and/or merge nodes so each belongs to a single temporal path.

@@

\subsection{Overview of the prototype interface}
\label{sec:ui}
To validate and explore the results of our methodology, we built a user
interface, illustrated in Figure~\ref{fig:ui}, considering census tract (CT)
level data from the Chicago region between 1970 and 2010. This region is known
for its entrenched racial divide and the emergence of a \emph{'young urban'}
population with a higher education level~\citep{Delmelle2016,Delmelle2017}. More
details are presented in Section~\ref{sec:study}.

\begin{figure*}
    \centering 
    \includegraphics[width=0.975\linewidth]{ui.pdf}
    \caption{Initial interface of our method showing the demographic evolution of Chicago and identifying the objectives of each plot.
        \textbf{(what)}: Cluster overview illustrating the most relevant aspect for each cluster. 
        \textbf{(where)}: Maps illustrating the geographical location, both as an overall summary and for each year.
        \textbf{(how)}: Tabular and visual summmarisation of how the regions classification changed over time.
        \textbf{(who)}: Demographic details for the currently selected data.
        \label{fig:ui}}
\end{figure*}

As illustrated by Figure~\ref{fig:ui}\unsure{CHECK image/ sec 4.5}, our proposed
interface heavily relies on colour to express cluster-related information. We
adopted this convention because colours can be used in all our visual tools in a
coherent manner. However, there is a limit on the number of distinct colours
that can be used. We limited the number of clusters to eight because this was
the largest number of colours that we could reliably and accessibly use, derived
from the 8-class Dark2 set from ColorBrewer~\citep{ColorBrewer}.

The configuration panel, on top left in Figure~\ref{fig:ui}, displays the
processing configuration of the visualised results: which aspects were used and
their weights (following Equation~\ref{eq:dist}). It also includes other
configuration options that can be altered without re-processing the data, such
as the number of clusters and the colour option. The gear button allows access
to the other configuration options that do require further processing, such as
changing location, aspects, and weights.

The cluster overview panel, marked as \emph{what} in Figure~\ref{fig:ui},
displays the most relevant aspects for each cluster, based on the distance
between the IQRs, as detailed in Section~\ref{sec:relevance}.\revision{2.5}{This
feature allows the user to quickly understand what exists in this region and
which aspect makes each cluster unique, without going into excessive detail. In
Chicago's example, we expect race to be relevant in general, and indeed, it is
recognised as the most relevant aspect for three of the four most distinct
clusters, with the fourth being education level. As we increase the number of
clusters, they get more nuanced, requiring more than one aspect for their proper
characterisation, as each cluster is divided into its composing sub-clusters.
The \emph{View all} button opens a new panel where all aspects are included,
while the chevron at the side lets the user expand each cluster separately,
allowing the exploration of more details on demand.}

@@

With the knowledge of what clusters are present, we can explore where they are
through the maps, marked as \emph{where} in Figure~\ref{fig:ui}. While the maps
on the lower part of the interface show the location of the clusters at each
time, the larger map shows a summarised version, based on the trajectories of
each region. \revision{1.9}{In simplified mode the colour corresponds to the
cluster present in that trajectory for more than half the temporal samples, and
is grey otherwise. With simplified colours disabled, the colours are the average
of the colours of the clusters in that trajectory in RGB colour
space~\footnote{\url{https://gka.github.io/chroma.js/\#chroma-average}}.}

\revision{2.5}{Each trajectory represents a different temporal path through the
clusters. Therefore the trajectories encode how the demographic characteristics
changed over time. These changes are summarised by the two sections identified
as \emph{how} in Figure~\ref{fig:ui}. In the left panel, there is a Sankey
diagram where each ribbon represents a different trajectory. The widths are
proportional to the population involved.} In our example in Figure~\ref{fig:ui},
the orange and green clusters contain most of the population and are fairly
stable over time. The pink cluster is small and mostly stable. The purple
cluster is increasing, mostly by incorporating areas that were previously
orange. Since the purple group corresponds to the emergent 'young urban' group,
this corroborates the findings of Delmelle~\citep{Delmelle2016,Delmelle2017},
showing that our network-based method can recover results from the traditional
data processing approach.

\revision{2.5}{In the next panel, illustrated in the top middle of
Figure~\ref{fig:ui}, is a transition matrix between the clusters. It indicates a
rounded percentage of regions whose area changed between each pair of
clusters. In this example, we can see that all clusters are fairly stable, with
more than ninety percent of identity transitions.} This kind of table can be
found in the related literature~\citep{Delmelle2016}, so it is familiar to the
advanced users. It not only informs the proportional changes, but allows the
selection of specific changes for further analysis.

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, marked as \emph{who}
in Figure~\ref{fig:ui}. The stacked bar plots summarise the overall demographic
composition of these regions. In this example, the maps show the transition from
orange to green and purple in several regions over time. Each aspect is
represented by a stacked bar plot, where the width of each rectangle corresponds
to \revision{1.10}{the percentage of population in that variable over the
considered period. In this case, a little less than half of the population are
married, and the percentage that are Widowers or Divorced is roughly similar.
About half of the population work in Administrative jobs, a third never
completed high-school, approximately forty percent have gross family income
below 50,000USD per year. About sixty percent identify as white. Placing the
mouse over one of the bars will open a small panel with the temporal evolution
of the population percentage of that specific variable, and clicking on the
chevron on the right side expands the corresponding aspect, showing census tract
level statistics, with details of the temporal evolution of each variable and
also the corresponding IQRs for the whole city. Those values differ from the
total population percentage, while the overall population of this region is
about sixty percent White, the average percentage of White population over the
CTs is a little over forty percent.}

More importantly, this interface is fully interactive, allowing for a
progressive exploration of the data. Clicking on a cluster in the cluster
overview panel and the other panels will filter the results to consider only
that cluster, highlighting the geographic regions and involved changes, and
updating population count and demographic composition. If the user selects a
region from the larger map, all identical trajectories are considered, and the
whole interface changes to allow for further study of that specific region.
Clicking on a region in small maps will bring up a new panel with the original
census data of this specific region. Further, these selectors can be combined in
different ways, enabling complex queries with instantaneous results.
